\section{Downloading \RSAT from the CVS server}

\textbf{Warning:} the CVS distribution is reserved to the members of
the \RSAT team. For external users, \RSAT is distributed as a tar
archive (file \file{rsat\_YYYY-MM-DD.tar.gz}, where YYYY-MM-DD
indicates the release date). If you are not a member of the \RSAT team
you can thus skip this section.

\subsection{CVS configuration}

You need to indicate your \program{cvs} client to use \program{ssh}
as remote shell application. For this, you can specify an environment
variable.

if your shell is tcsh, add the following line to the \file{.cshrc}
file in your home directory.

\begin{lstlisting}
setenv CVS_RSH ssh
\end{lstlisting}


If your shell is bash, add the following line to the \file{.bashrc}
file in your home directory.

\begin{lstlisting}
export CVS_RSH=ssh
\end{lstlisting}

\subsection{Obtaining a first version of \RSAT distribution}

The following command should be used the first time you retrieve the
tools from the server (you need to replace \texttt{[mylogin]} by the
login name you received when signing the \RSAT license).

\begin{lstlisting}
cvs -d [mylogin]@cvs.bigre.ulb.ac.be:/cvs/rsat co rsat
\end{lstlisting}


This will create a directory \file{rsat} on your computer, and
store the programs in it. Note that at this stage the programs are not
yet functional, because you still need to install genomes, which are
not included in the CVS distribution.

\subsection{Updating \RSAT programs}

Once the tools have been retrieved, you can obtain updates very
easily. For this, you need to change your directory to the rsat
directory, and use the \texttt{cvs} command in the following way.

\begin{lstlisting}
cd rsat
cvs -q update -d
\end{lstlisting}


