\chapter{Pathway inference}

\section{Introduction}
The idea of pathway inference is to connect a given set of seed nodes
in the network and thereby extracting a sub-network that is optimal
according to certain criteria (e.g. minimal weight or maximal relevance).\\
In the context of biological networks, the goal is to obtain a valid pathway for a set
of biological entities of interest, e.g. genes from microarray data or compounds from metabolomic data.
For instance, genes whose products participate in the same metabolic pathway are often co-expressed or grouped
together in operons or regulons. We may try to reconstruct this metabolic pathway by associating
the gene products to relevant reactions and connecting these reactions in a metabolic network.
The resulting sub-network may be a known metabolic pathway or an unknown pathway consisting of known
pathways or known reactions and compounds. In the context of microarray data, pathway inference from a
set of co-expressed genes may predict which pathways are up- or down-regulated.\\

\section{Inferring a pathway for a set of co-expressed genes}

As an example, we take the case study discussed in \cite{vanHelden01}.
In this case study, a pathway is assembled from genes in the cell-cycle regulated MET cluster \cite{Spellman98}.
Results described in this tutorial have been obtained with KEGG RPAIR version 49.0.

\subsection{Protocol for the web server}

\begin{enumerate}

\item In the \neat  menu, select the entry \program{Pathwayinference}.

\item Copy-paste the gene names below in the seed nodes text field:\\
Met3\\
Met14\\
Met16\\
Met5\\
Met10\\
Met17\\
Met6

\item Select "Genes/Enzymes" as identifier type.

\item In the text field "Genes/Enzymes are from organism" type sce, the KEGG abbreviation for \textit{Saccharomyces cerevisiae}.

\item Push the \option{GO} button.

\end{enumerate}

The result of the mapping of the given genes to KEGG RPAIRS (reactant pairs, \cite{Kotera2004}) is displayed.
Since more than one reactant pair is associated to each gene, we end up with a group of reactant pair groups.
Note that each gene (except for Met5) is associated to one or more EC numbers, each of which has been mapped
to its corresponding reactions in KEGG, which have in turn be mapped to their corresponding reactant pairs.\\

You can now select how to deal with the groups. This is a sensitive choice that strongly
affects the inferred pathway and which depends on your data.
In case of reactant pairs, it is usually not recommended to keep the groups, because
they might contain mutually exclusive reactant pairs (that is reactant pairs belonging to the same reaction).
We prevent pathwayinference in these cases, to avoid the same reaction occurring more than once in the resulting sub-network.
Instead, the groups can be re-arranged such that mutually exclusive reactant pairs are joined.
The original groups in this case are replaced by groups that reflect
"possible catalytic mechanisms" enabled by the given enzymes.\\
In general, if you keep the original groups, you assume implicitely that only
a subset of the reactions associated to the given gene will be active in the pathway.
If you think that all reactions associated to a gene might be active, choose "Treat each group member as a separate group." or,
in case of reactant pairs, choose "Treat each group member as a separate group, but ignore group members that occur more than once and merge exclusive reactant pairs into groups.".\\

For the study case, we recommend you to choose "Treat each group member as a separate group, but ignore group members that occur more than once and merge exclusive reactant pairs into groups.".

Push \option{GO}. In a few minutes, the result page will be displayed.


\subsection{Protocol for the command-line tools}

This section assumes that you have installed the RSAT/NeAT command line tools.

Pathwayinference is a web application that calls the pathwayinference web service.
You can use the Pathwayinference command line tool on the networks provided in the
network repository (check the Pathwayinference Manual for this) to reproduce
results obtained with the web application on command line. Note that the mapping of genes to reactions
and re-grouping of seeds can only be done via the web application.

Type the following command in one line:
{\color{Blue} \begin{footnotesize}
		\begin{verbatim}
java -Xmx800m graphtools.algorithms.Pathwayinference -g RPAIRGraph_allRPAIRs_undirected.txt -f flat
	     -s 'RP00016#RP00182/RP00647/RP00561/RP00143#RP00960#RP04049/RP00096#RP00168#
	     RP04532/RP00003/RP00446/RP00946#RP00857/RP04474/RP00050#RP04533'
	     -b -y con -P -u -x 0.05
	\end{verbatim} \end{footnotesize}
	}

\subsection{Interpretation of the results}

The resulting sub-network contains the pathway given in \cite{vanHelden01}. The reaction catalysed by
the Met5 enzyme (which was not found in KEGG) has been correctly inferred.

The pathway described in the study case unites the sulfur assimilation and methionine biosynthesis pathways. It consists of the following steps: \\
Sulfate 2.7.7.4 Adenylyl sulfate 2.7.1.25 3'phosphoadenylylsulfate 1.8.99.4 sulfite 1.8.1.2 sulfide (alias hydrogen sulfide) 4.2.99.10 Homocysteine 2.1.1.14 L-Methionine\\

The matching part of the inferred pathway is:\\
\textbf{RP00960} Adenylyl sulfate RP05930 Sulfite \textbf{RP00168} Hydrogen sulfide RP01406 L-Homocysteine \textbf{RP0096}\\
Seeds are printed in bold.

In addition, the inferred pathway contains a branch that leads
from Adenylylselenate via 3'-Phosphoadenylylselenate, Selenite and hydrogen selenide to selenocysteine.
This branch mirrors sulfur incorporation, but instead of sulfur, selenium is incorporated into Homocysteine.\\

The presence of both the selenium and sulfur incorporation pathways in the inferred sub-network
reflects the well-known fact that selenium might replace sulfur in metabolism.\\

This example demonstrated that given a set of differentially expressed genes from
micro-array data and a metabolic network, it is possible to infer a metabolic pathway
that might be affected by altered expression of the query genes.

\section{Summary}

Pathwayinference allows extraction of sub-networks from larger networks given a set of seed nodes. The web application
is tailored to metabolic networks, but non-metabolic networks can be processed as well.

\section{Strengths and Weaknesses of the approach}

\subsection{Strengths}

\begin{enumerate}
\item Sub-network extraction can be applied to any biological network.

\item It can discover unknown pathways consisting of known components.

\item It can be fine-tuned to favor certain nodes. For instance, in a global metabolic network,
      reactions/compounds known to occur in certain species might receive a weight much lower than other nodes,
      to favor extraction of species-specific sub-networks.

\item Groups of seed nodes can be specified to reflect AND/OR relationships between seeds.

\item The web application allows metabolic pathway inference given compounds, reactions, reactant pairs, EC numbers or gene identifiers
      as seeds in metabolic networks extracted from the two major metabolic databases KEGG \cite{Kanehisa2008} and MetaCyc \cite{biocyc}.

\item Metabolic sub-network extraction has been validated on 71 metabolic pathways extracted from MetaCyc.
\end{enumerate}

\subsection{Weaknesses}

\begin{enumerate}

\item In general, the accuracy of pathway inference depends on the quality of the given network and the number of seeds available.

\item It is not possible to infer other than trivial cycles. Consequently, cyclic and spiral-shaped metabolic pathways can only be partly
inferred.

\item In the densely connected region of metabolic networks, metabolic pathway inference cannot well distinguish alternative pathways
without a large number of seed nodes.

\item The algorithms are too time-consuming to estimate p-values by computing a score distribution (where the score would be the sub-network weight)
for randomly chosen seed nodes on the fly. We envisage to pre-compute these distributions for the pre-loaded networks.

\end{enumerate}

\section{Troubleshooting}

\begin{enumerate}

\item Pathwayinference parameter error.

	You provided insufficient or invalid parameters. Please check the pathwayinference manual page.

\item You did not specify enough valid seed node groups! Pathwayinference needs at least two valid seed node groups.

	For the pre-loaded metabolic networks from KEGG and MetaCyc, each seed
	is mapped to data (e.g. compound/reaction identifiers, EC numbers) from these two databases.
	If the seeds do not map anything, they are considered to be invalid.
	At least two valid seed groups are needed to infer a network.

\item The node with identifier ID is not part of the input graph.

	Make sure that your input network contains the node with given identifier.

\item There are seeds that exclude each other mutually between groups!

	Your seed nodes have been mapped to reactant pairs belonging to different groups. At least one of these groups contains
	a reactant pair that is mutual exclusive to a reactant pair in another group (that is both reactant pairs belong to the
	same reaction). You may choose another group treatment, as for example:
	"Keep groups but merge overlapping groups and groups with inter-group exclusive reactant pairs.".

\item Pathwayinference failed to extract a subgraph.

    None of the seed node groups could be connected to any other seed node group. Each might belong to a
    separate component of the input network.

\end{enumerate}