%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliographic references
\chapter{References}

\begin{enumerate}

\item van Helden, J., Andre, B. \& Collado-Vides,
  J. (1998). Extracting regulatory sites from the upstream region of
  yeast genes by computational analysis of oligonucleotide
  frequencies. J Mol Biol 281(5), 827-42.

\item van Helden, J., André, B. \& Collado-Vides, J. (2000). A web
  site for the computational analysis of yeast regulatory
  sequences. Yeast 16(2), 177-187.

\item van Helden, J., Olmo, M. \& Perez-Ortin,
  J. E. (2000). Statistical analysis of yeast genomic downstream
  sequences reveals putative polyadenylation signals. Nucleic Acids
  Res 28(4), 1000-1010.

\item van Helden, J., Rios, A. F. \& Collado-Vides,
  J. (2000). Discovering regulatory elements in non-coding sequences
  by analysis of spaced dyads. Nucleic Acids Res. 28(8):1808-18.

\item van Helden, J., Gilbert, D., Wernisch, L., Schroeder, M. \&
  Wodak, S. (2001). Applications of regulatory sequence analysis and
  metabolic network analysis to the interpretation of gene expression
  data. Lecture Notes in Computer Sciences 2066: 155-172.

\item van Helden, J. 2003. Prediction of transcriptional regulation by
  analysis of the non-coding genome. Current Genomics 4: 217-224.

\item van Helden, J. 2003. Regulatory sequence analysis tools. Nucleic
  Acids Res 31: 3593-3596.

\item van Helden, J. 2004. Metrics for comparing regulatory sequences
  on the basis of pattern counts. Bioinformatics 20: 399-406.

\item Simonis, N., J. van Helden, G.N. Cohen, and
  S.J. Wodak. 2004. Transcriptional regulation of protein complexes in
  yeast. Genome Biol 5: R33.

\item Simonis, N., S.J. Wodak, G.N. Cohen, and J. van
  Helden. 2004. Combining pattern discovery and discriminant analysis
  to predict gene co-regulation. Bioinformatics.

\end{enumerate}

