%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% MARKOV MODELS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Markov models}

Markov models allow to represent local dependencies between successive
residues. A Markov model of order $m$ assumes that the probability to
find the residue $r$ at position $i$ of a sequence depends on the
$m$ preceding residues.

\subsection{Transition frequency tables}

Markov models are described by transition frequencies $P(R|W_m)$,
i.e. the probability to osberve residue $R$ at a certain position,
depending on the preceding word $W_m$ of size $m$.

\subsection{Oligonucleotide frequency tables}

\RSAT allows to derive organism-specific Markov models from
oligonucleotide frequency tables.

Pre-calibrated oligonucleotide frequency tables are stored in the form
of oligonucleotide frequency tables (see chapter on pattern
discovery).

The calibration tables for \org{Escherichia coli K12} can be found in
the \RSAT directory \file{oligo-frequencies}.

\begin{small}
\begin{verbatim}
cd $RSAT/data/genomes/Escherichia\_coli\_K12/oligo-frequencies
ls -ltr
\end{verbatim}
\end{small}

For example, the file
\file{4nt\_upstream-noorf\_Escherichia\_coli\_K12-1str.freq.gz}
indicates the tetranucleotide frequencies for all the upstream
sequences of \org{E.coli}.

\begin{small}
\begin{verbatim}
cd $RSAT/data/genomes/Escherichia_coli_K12/oligo-frequencies/

## Have a look at the content of the 4nt frequency file
gunzip -c 4nt_upstream-noorf_Escherichia_coli_K12-1str.freq.gz | more
\end{verbatim}
\end{small}

\subsection{Converting oligonucleotide frequencies into transition frequencies}

Transition frequencies are automatically derived from the table of
oligonucleotide frequencies, but one should take care of the fact
that, in order to estimate the transition frequencies for a Markov
model of order $m$, we need to use the frequency tables for
oligonucleotides of size $m+1$.

We can illustrate this by converting the table of dinucleotide
frequencies into a transition matrix of first order. For this, we can
use the program \program{convert-background-model}.

\begin{small}
\begin{verbatim}
convert-background-model \
  -i 2nt_upstream-noorf_Escherichia_coli_K12-1str.freq.gz  \
  -from oligo-analysis -to tab
\end{verbatim}
\end{small}

The output displays the transition matrix of a Markov model of order
1.  Each row of the transition matrix indicates the prefix $W_m$, and
each column the suffix $r$. For a Markov model of order 1, the
prefixes are single residues.

We can now calculate a Markov model of 2nd order, from the table of
trinucleotide frequencies.

\begin{small}
\begin{verbatim}
convert-background-model \
  -i 3nt_upstream-noorf_Escherichia_coli_K12-1str.freq.gz  \
  -from oligo-analysis -to tab
\end{verbatim}
\end{small}

The transition matrix contains 16 rows (prefixes, corresponding to
dinucleotides) and 4 columns (the suffixes, corresponding to
nucleotides).

The same operation can be extended to higher order markov models.

\subsection{Bernoulli models}

In contrast with Markov model, Bernoulli models assume that the
residue probabilities are independent from the position. By extension
of the concept of Markov order, Bernoulli models can be conceived as a
Markov model of order 0. We can thus derive a Bernoulli model ($m=0$)
from the nucleotide frequencies ($m+1=1$).

\begin{small}
\begin{verbatim}
convert-background-model \
  -i 1nt_upstream-noorf_Escherichia_coli_K12-1str.freq.gz  \
  -from oligo-analysis -to tab
\end{verbatim}
\end{small}

The suffix column is now empty (there is no suffix, since the order is
0), and the matrix simply displays 4 columns with the frequencies of
\seq{A}, \seq{C}, \seq{G} and \seq{T}.
