%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% PATHWAY EXTRACTION TOOLS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Pathway extraction tools}

\section{Using pathway extraction tools}

\subsection{Listing tools and getting help}
You can list available tools by typing:

\begin{lstlisting}
java graphtools.util.ListTools
\end{lstlisting}

All tools provide a \option{-h} option to display help.

\subsection{Abbreviating tool names}
The command line tool names may be simplified by setting aliases.
For example, in the bash shell:

\begin{lstlisting}
alias Pathfinder="java graphtools.algorithms.Pathfinder"
\end{lstlisting}

allows to type:

\begin{lstlisting}
Pathfinder -h
\end{lstlisting}

instead of:

\begin{lstlisting}
java graphtools.algorithms.Pathfinder -h
\end{lstlisting}

\subsection{Increasing JVM memory}
For large graphs, you may need to increase the memory allocated to the java virtual machine.
You can do so by specifying the \option{-Xmx} option.

Example:

\begin{lstlisting}
java -Xmx800m graphtools.algorithms.Pathfinder -h
\end{lstlisting}

\section{Obtaining metabolic networks}

\subsection{Downloading MetaCyc and KEGG generic metabolic networks from the
NeAT web server}\label{Download}

Metabolic networks can be downloaded from the NeAT web server.
Go to the menu entry ``Path finding and pathway extraction'', open the
``Pathway extraction'' page and click on ``More networks can be downloaded
here.'' This will open a table with tab-delimited generic MetaCyc and KEGG
networks.

\subsection{Building KEGG generic metabolic networks} 

\minisec{Reaction network}
To build the directed reaction network, type:

\begin{lstlisting}
java -Xmx800m graphtools.parser.KeggLigandDataManager -m
\end{lstlisting}

The network is stored in the current directory.

The execution of this command takes quite long, because it fetches the reaction
and compound files from KEGG's ftp repository at \url{ftp.genome.jp}. 
To get these files, the \program{KeggLigandDataManager} requires
\program{wget} to be installed and in your path. \program{wget} is 
freely available from \url{http://www.gnu.org/software/wget/}.

Alternatively, you may first download the reaction and compound files yourself
from the KEGG ftp server. Type in your browser (or in your favourite ftp
client):\\
\url{ftp://anonymous@ftp.genome.jp/pub/kegg/ligand/compound/compound}\\
and save the compound file into \file{\$RSAT/data/KEGG/KEGG\_LIGAND}. Do
the same for the reaction file at\\
\url{ftp://anonymous@ftp.genome.jp/pub/kegg/ligand/reaction/reaction}.\\ 
Then you can run the command above to generate the reaction network.

\minisec{RPAIR network}\label{RPAIR}
To construct the undirected RPAIR network, type:

\begin{lstlisting}
java -Xmx800m graphtools.parser.KeggLigandDataManager -s -u
\end{lstlisting}

Creating the RPAIR network will also create the \file{rpairs.tab} file, which
can be placed in the KEGG directory for later use by typing:

\begin{lstlisting}
cp $RSAT/data/KEGG/KEGG_LIGAND/rpairs.tab $RSAT/data/KEGG/rpairs.tab
\end{lstlisting}

An older version of this file is also available from the \neat web server
in the data/KEGG directory.

\minisec{Reaction-specific RPAIR network}
For the reaction-specific undirected RPAIR network, type:

\begin{lstlisting}
java -Xmx800m graphtools.parser.KeggLigandDataManager -t -u
\end{lstlisting}

\subsection{Building KEGG organism-specific metabolic networks}

The MetabolicGraphProvider tool allows you to merge KEGG KGML files into
a metabolic network specific to a set of organisms. 

\minisec{Prerequisites}
You may first create the list of available KEGG organisms:

\begin{lstlisting}
java -Xmx800m graphtools.parser.MetabolicGraphProvider -O
\end{lstlisting}

This command will create the file \file{Kegg\_organisms\_list.txt} in the
current directory. Since this file is needed by the
\program{MetabolicGraphProvider}, you may copy it to its default location:

\begin{lstlisting}
cp Kegg_organisms_list.txt \$RSAT/data/KEGG/Kegg_organisms_list.txt
\end{lstlisting}

Alternatively, you may obtain an older version of this file from the \neat web
server in the data/KEGG directory. 

\minisec{Creating an organism-specific reaction network for E. coli}

The command below builds the \org{E. coli}-specific metabolic reaction network
from its KGML files:

\begin{lstlisting}
java -Xmx800m graphtools.util.MetabolicGraphProvider -i eco -o ecoNetwork.tab
\end{lstlisting}

The KGML files are automatically obtained from the current KEGG database
(which may take very long). Alternatively, they can be downloaded manually from
\url{http://www.genome.jp/kegg/xml/}.
If downloaded manually, all organism-specific KGML files have to be placed in a
folder named with the organism's KEGG abbreviation (e.g.\textit{eco}
for \org{E. coli}). The folder should be located in the \file{\$RSAT/data/KEGG}
directory.
 
We can also merge the KGML files of several organisms into one network and apply
some filtering as follows (in one line):

\begin{lstlisting}
java -Xmx800m graphtools.util.MetabolicGraphProvider -i ecv/eco -o
eco_ecv_Network.tab -c C00001/C000002/C00003/C00004/C00005/C00006/C00007/C00008
\end{lstlisting}

This command will construct a merged metabolic network from two
\org{E. coli} strains (\org{Escherichia coli K-12 MG1655} and \org{Escherichia
coli O1 (APEC)}) 
and in addition filter out some highly connected compounds 
(water, ATP, NAD+, NADH, NADPH, NADP+, oxygen and ADP).

\subsection{Building metabolic networks from biopax files}

Several metabolic databases store their data in biopax format
(\url{http://www.biopax.org/}), e.g. BioCyc and Reactome. You can create
a metabolic network from a biopax file using the
\program{GDLConverter}.

For instance, you may download the lysine biosynthesis I pathway from
\url{http://metacyc.org/} in biopax format and save it into a file named
\file{lysine\_pwy1.xml}. You can then obtain a tab-delimited metabolic network
from this file using the command below (in one line). Note that the metabolic
network preserves the reaction directions indicated in the biopax file, that is irreversible reactions are
kept.

\begin{lstlisting}
java graphtools.util.GDLConverter -i lysine_pwy1.xml 
-o lysine.txt -O tab -I biopax -b -d
\end{lstlisting}

Option \option{-O} indicates the output format (tab-delimited), \option{-I}
specifies the input format (biopax in this case), \option{-b} flags that
attributes required for the metabolic format should be set and \option{-d} tells
the program to construct a directed network.

The \program{GDLConverter} may be applied in general to interconvert networks
in different formats.

\section{Finding k-shortest paths}

Pathways may be extracted from metabolic networks by enumerating the
\textit{k}-shortest paths between a set of source compounds/reactions and a
set of target compounds/reactions.

In metabolic networks, some compounds such as ATP or NADPH are involved in a
large number of reactions, thus acting as shortcuts for the path finding
algorithm. However, paths crossing these highly connected compounds are not
biochemical relevant. In order to prevent the path finding algorithm to traverse
these compounds, the metabolic network should be weighted.

For example, assume you have generated (\ref{RPAIR}) or downloaded
(\ref{Download}) a KEGG RPAIR network stored in the file 
\file{KEGG\_RPAIR\_undirected.txt}. Given this network, we can list the 
three highest-ranked lightest paths between
aspartate (KEGG identifier: C00049) and lysine (KEGG identifier: C00047) with
the command below (in one line):

\begin{lstlisting}
java -Xmx800m graphtools.algorithms.Pathfinder -g KEGG_RPAIR_undirected.txt 
-s C00049 -t C00047 -y con -b -r 3 -f tab
\end{lstlisting}

where option \option{-s} specifies the source node (more than one can be given),
\option{-t} the target node (as for the source, more than one target can be
specified),\option{-f} indicates the format of the input network
(tab-delimited), \option{-r} indicates the rank, option{-y} gives the weight policy to be applied
(con sets the weight of compounds to their degree and the weight of reactions
to one) and \option{-b} flags that the input network is metabolic.

This command will yield the following output (with KEGG RPAIR version 49.0):

\begin{lstlisting}
INFO: Pathfinder took 5014 ms to perform its task.
; Experiment exp_0
; Pathfinding results
; Date=Fri Apr 30 16:34:27 CEST 2010
; ===============================
; INPUT
; Source=[C00049]
; Target=[C00047]
; Graph=KEGG_RPAIR_undirected.txt
; Directed=false
; Metabolic=true
; RPAIR graph=true
; CONFIGURATION
; Algorithm=rea
; Weight Policy=con
; Maximal weight=2147483647
; Exclusion attribute=ExclusionAttribute
; Rank=3
; REA timeout in minutes=5
; EXPLANATION OF COLUMNS
; Start node=given start node identifier
; End node=given end node identifier
; Path=path index
; Rank=rank of path (paths having same weight have 
the same rank, though their step number might differ)
; Weight=weight of path (sum of edge weights)
; Steps=number of nodes in path
; Path=sequence of nodes from start to end node that forms the path
; ===============================
#start  end     path    rank    weight  steps   path
C00049  C00047  1       1       122.0   15      C00049->RP00932->C03082
->RP02107->C00441->RP02109->C03340->RP00740->C03972->RP03970->C03871
->RP02474->C00680->RP00907->C00047
C00049  C00047  2       2       126.0   15      C00049->RP00932->C03082
->RP02107->C00441->RP02109->C03340->RP00740->C03972->RP11205->C00666
->RP02449->C00680->RP00907->C00047
C00049  C00047  3       3       134.0   11      C00049->RP00116->C00152
->RP06538->C00151->RP01393->C00405->RP07206->C00739->RP00911->C00047
C00049  C00047  4       4       143.0   13      C00049->RP03035->C04540
->RP01395->C00152->RP06538->C00151->RP01393->C00405->RP07206->C00739
->RP00911->C00047
\end{lstlisting}

The format of the output can be changed to output the path list as a network.
This network can then be visualized using the \program{PathwayDisplayer} as
explained in section \ref{Visualize}.

To output the path list as a network in gml format, run the following command
(in one line):

\begin{lstlisting}
 java -Xmx800m graphtools.algorithms.Pathfinder -g KEGG_RPAIR_undirected.txt 
 -s C00049 -t C00047 -y con -b -r 3 -f tab -T pathsUnion -O gml 
 -o asp_lys_paths.gml
\end{lstlisting}

The file \file{asp\_lys\_paths.gml} created in the current directory contains
the network in gml format.

\section{Linking genes to reactions}

The main application of pathway extraction is to interpret a set of
associated enzyme-coding genes. An association can for example be co-expression
in a microarray, co-regulation in an operon or regulon or co-occurrence in a
phylogenetic profile.

In this section, we will see how to link enzyme-coding genes to their reactions.
This is not a straightforward task, as an N:N relationship exists between genes,
EC numbers, reactions and reactant pairs.

\subsection{Prerequisites}\label{metabolicdb}
In order to link genes to reactions, the metabolic database needs to be
installed. The installation of this database is described in chapter
``Metabolic Pathfinder and Pathway extraction" in the \neat web server install
guide, which is available from the \neat web server download section. 

\subsection{Linking genes of the isoleucine-valine operon to
reactions}\label{ilv_operon}

The isoleucine-valine operon (RegulonDB identifier: ilvLG\_1G\_2MEDA) in
\org{Escherichia coli} is known to contain enzymes of
the isoleucine and valine biosynthesis pathway. 

It consists of the following genes:

\begin{lstlisting}
ilvL ilvG_1 ilvG_2 ilvM ilvE ilvD ilvA
\end{lstlisting}

These genes can be linked to KEGG reactant pairs using the command below (in one
line):

\begin{lstlisting}
java graphtools.util.SeedConverter -i ilvL/ilvG_1/ilvG_2/ilvM/ilvE/ilvD/ilvA 
-I string -O eco -o ilv_operon_seeds.txt -r
\end{lstlisting}

Option \option{-r} flags that genes should be mapped to (main) reactant pairs,
\option{-O} specifies the source organism of the genes, \option{-i} lists the
genes and \option{-I} specifies the input format.

\section{Predicting metabolic pathways}

Given a set of seeds (compounds or reactions/reactant pairs) and a metabolic
network, the task of the pathway extraction tool is to extract a metabolic
pathway that connects these seeds in the metabolic network. The tool is quite generic and can be
applied to any network and seed node set. However, it has been tailored to
metabolic pathway prediction.

\subsection{Predicting a metabolic pathway for the isoleucine-valine operon}

Assume you have generated the seed input file from section \ref{ilv_operon} and
the KEGG RPAIR graph as described in section \ref{RPAIR}. The KEGG RPAIR graph
is assumed to be stored in a tab-delimited file named
\file{KEGG\_RPAIR\_undirected.txt}. Then we can predict the pathway for the
genes in the isoleucine-valine operon with the following command (in one line):

\begin{lstlisting}
java -Xmx800m graphtools.algorithms.Pathwayinference -g
KEGG_RPAIR_undirected.txt -i ilv_operon_seeds.txt -b -f tab  
-y con -E Result -a takahashihybrid -U -o ilv_predicted_pathway.tab
\end{lstlisting}

where option \option{-b} specifies that the network is a metabolic
network, \option{-f} indicates the input network format (tab-delimited),
\option{-a} specifies the algorithm to be used and \option{-y} indicates the
weight policy to be applied (con stands for connectivity, which means that
compound nodes receive a weight corresponding to their degree). 
Option \option{-E} is used to indicate the name of the folder where results are
stored. This is especially useful when several predictions are carried out in a
row, because the output file in this case reports the merged pathway. In the
example above, the result folder serves to store the properties of
the predicted pathway (obtained with option \option{-U}).

A variant of the pathway extraction exploits the fact that we work with the
KEGG RPAIR graph, which allows us to link adjacent main reactant pairs (i.e.
reactant pairs sharing a compound). This is done in a preprocessing step
(option \option{-P}):

\begin{lstlisting}
java -Xmx800m graphtools.algorithms.Pathwayinference -g
KEGG_RPAIR_undirected.txt -i ilv_operon_seeds.txt -b -f tab -y con -P 
-a takahashihybrid -o ilv_predicted_pathway_preprocessed.tab
\end{lstlisting}

\subsection{Mapping reference pathways onto the predicted pathway}

The predicted metabolic pathway can be mapped to reference pathways stored in
the metabolic database. This can be done as follows:

\begin{lstlisting}
java graphtools.util.MetabolicPathwayProvider -i ilv_predicted_pathway.tab 
-I tab -D KEGG -o ilv_predicted_pathway_mapped.tab
\end{lstlisting}

where option \option{-D} indicates that reference pathways should be taken from
KEGG and \option{-I} indicates the input format of the pathway. In the output
pathway, nodes mapping to reference pathways are annotated with
a color and the name of the corresponding reference pathway. 
The program  also outputs the color-code of mapping reference pathways:

\begin{lstlisting}
INFO: Legend
BurlyWood: Valine,_leucine_and_isoleucine_biosynthesis
orange: no match to any reference pathway
\end{lstlisting}

\subsection{Annotating the predicted pathway}

The nodes of a predicted metabolic pathway can be labeled with names
(compounds), EC numbers (reactions) and genes (reactions). 
The requires the metabolic database to be installed (see
\ref{metabolicdb}).

The command below annotates the metabolic pathway named
\file{ilv\_predicted\_pathway.tab} and colors its seed
nodes (stored in the seed node file \file{ilv\_operon\_seeds.txt} ) in blue:

\begin{lstlisting}
java graphtools.util.GraphAnnotator -i ilv_predicted_pathway.tab -I tab 
-o ilv_predicted_pathway_annotated.tab -O tab -k -b 
-F ilv_operon_seeds.txt
\end{lstlisting}

Option \option{-k} tells \program{GraphAnnotator} to associate EC numbers to
KEGG genes using the current KEGG database, {-b} indicates that the pathway is a metabolic
pathway, {-I} specifies the input format of the pathway to be annotated
(tab-delimited) and {-F} indicates the location of the seed node file.

\subsection{Visualizing the predicted pathway}\label{Visualize}

The visualization of a pathway requires \program{graphviz} to be installed,
which is available here \url{http://www.graphviz.org/}.

With \program{graphviz} installed, the pathway can be visualized as follows:

\begin{lstlisting}
java graphtools.util.PathwayDisplayer -i ilv_predicted_pathway_annotated.tab 
-I tab -p
\end{lstlisting}

Option \option{-p} tells \program{PathwayDisplayer} to generate the image with
\program{graphviz}, \option{-I} indicates the input format of the pathway to be
displayed (tab-delimited).

