%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% PATTERN DISCOVERY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Matrix-based Pattern discovery}

\RSAT does not (yet) contain programs for matrix-based pattern
discovery. However, several excellent programs exist for matrix-based
pattern discovery, and it is often useful to combine various
approaches in order to compare the results and select the most
consistent ones. We show hereafter some examples of utilization for
some of these programs:

\begin{itemize}

\item \program{consensus}, a greedy approach of pattern discovery,
  developed by Jerry Hertz.

\end{itemize}

\section{consensus (program developed by Jerry Hertz)}

An alternative approach for matrix-based pattern discovery is
\textit{consensus}, a program written by Jerry hertz, an based on a
greedy algorithm. We will see how to extract a profile matrix from
upstream regions of the PHO genes.

\subsection{Getting help}

As for RSAT programs, there are two ways to get help from Jerry Hertz'
proigrams: a detailed manual can be obtained with the option
\option{-h}, and a summary of options with \option{-help}. Try these
options and read the manual.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
consensus -h
consensus -help
\end{verbatim} \end{footnotesize}
}


\subsection{Sequence conversion}


\textit{consensus} uses a custom sequence format. Fortunately, the RSAT
package contains a sequence conversion program (\textit{convert-seq})
which supports Jerry Hertz' format. We will thus start by converting
the fasta sequences in this format. 

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
convert-seq -i PHO_up800-noorf.fasta -from fasta -to wc -o PHO_up800-noorf.wc
\end{verbatim} \end{footnotesize}
}


\subsection{Running consensus}

Using consensus requires to choose the appropriate value for a series
of parameters. We found the following combination of parameters quite
efficient for discovering patterns in yeast upstream sequences.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
consensus -L 10 -f PHO_up800-noorf.wc -A a:t c:g -c2 -N 10
\end{verbatim} \end{footnotesize}
}

The main options used above are

\begin{description}
\item[-L 10] we guess that the pattern has a length of about 10 bp;

\item[-N 10] we expect about 10 occurrences in the sequence set. Since
  there are 5 genes in the family, this means that we expect on
  average 2 regulatory sites per gene, which is generally a good guess
  for yeast.

\item[-c2] indicates \program{consensus} that the motif can be searched
  on both strands.

\item[-A a:t c:g] specifies the alphabet. Indeed, \program{consensus}
  can be used to extract motif from DNA sequences, proteins, or a text
  based on an arbitrary alphabet. In thus tutorial we are only
  interested in DNA sequences, we wpecify thus \option{-A a:t c:g}
  (the semicolons indicate the complementary residues).
\end{description}

By default, several matrices are returned. Each matrix is followed by
the alignment of the sites on which it is based. Note that the 4
matrices are highly similar, basically they are all made of several
occurrences of the high afinity site \seq{CACGTG}, and matrices 1 and
3 contain one occurrence of the medium affinity site
\seq{CACGTT}. These matrices are thus redundant, and it is generally
appropriate to select the first one of the list for further analysis,
because it is the most significant matrix found by the program.

Also notice that these matrices are not made of exactly 10 sites
each. \textit{consensus} is able to adapt the number of sites in the
alignment in order to get the highest information content. The option
\option{-N 10} was an indication rather than a rigid requirement.

We can use the options \option{-pt 1} and \option{-pf 1} to restrict the
result to a single matrix (the most significant one). To save the
result in a file, we can use the symbol ``greater than'' ($>$) which
redirects the output of a program to a file.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
consensus -L 10 -f PHO_up800-noorf.wc -A a:t c:g -c2 -N 10 -pf 1 -pt 1 \
    > PHO_consensus_L10_N10_c2.matrix
\end{verbatim} \end{footnotesize}
}

(this may take a few minutes)

Once the task is achieved, check the result.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
more PHO_consensus_L10_N10_c2.matrix
\end{verbatim} \end{footnotesize}
}

\section{Random expectation}


{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
random-seq -format wc -r 10 -l 800 -bg upstream-noorf \
  -org Saccharomyces_cerevisiae -ol 6 -lw 0 -o rand_Sc_ol6_n10_l800.wc 

consensus -L 10 -f rand_Sc_ol6_n10_l800.wc -A a:t c:g -c2 -N 10 -pf 1 -pt 1 \
    > rand_Sc_ol6_n10_l800_L10_N10_c2.matrix
\end{verbatim} \end{footnotesize}
}
