%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% PATTERN DISCOVERY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{String-based pattern discovery}

In a pattern discovery problem, you start from a set of functionally
related sequences (e.g.  upstream sequences for a set of co-regulated
genes) and you try to extract motifs (e.g. regulatory elements) that
are characteristic of these sequences.

Several approaches exist, either string-based or
matrix-based. \concept{String-based pattern discovery} is based on an
analysis of the number of occurrences of all possibles words
(\program{oligo-analysis}), or spaced pairs
(\program{dyad-analysis}). The methods for \concept{matrix-based
  pattern discovery} rely on the utlisation of some machine-learning
method (e.g. greedy algorithm, expectation-maxiisation, gibbs
sampling, ...) in order to optimise of some scoring function
(log-likelihood, information,...) which is likely to return
significant motifs.

In this chapter we will mainly focus on string-based approaches, and
illustrate some of their advantages. A further chapter will be
dedicated to matrix-based pattern discovery.

For microbial cis-acting elements, string-based approaches give
excellent results. The main advantages of these methods:

\begin{itemize}
\item[+] Simple to use
\item[+] Deterministic (if you run it repeatedly, you always get the
  same result), in contrast with stochastic optimization methods.
\item[+] Exhaustive : each word or space pair is tested
  independently. Consequently, if a set of sequences contains several
  exceptional motifs, all of them can be detected in a single run. 
\item[+] The tests of significances can be performed on both tails of
  the theoretical distribution, in order to detect either
  over-represented, or under-represented patterns.
\item[+] Fast. 
\item[+] Able to return a negative answer: if no motif is significant,
  the programs return no motif at all. This is particularly important
  to reduce the rate of false positive.
\end{itemize}


An obvious advantage of matrix-based approach is that they provide a
more refined description of motifs presenting a high degree of
degeneracy. However, a general problem of matrix-based approaches is
that it is impossible to analyze all possible position-weight
matrices, and thus one has to use heuristics. There is thus a risk to
miss the global optimum because the program is attracted to local
maxima. Another problem is that there are more parameters to select
(typically, matrix width and expected number of occurrences of the
motif), and their choice drastically affects the quality of the
result.

Basically, I would tend to prefer string-based approaches for any
problem of pattern discovery. On the contrary, matrix-based approaches
are much more sensitive for pattern matching problems (see below). My
preference is thus to combine string-based pattern discovery and
matrix-based pattern matching.

But I am obviously biased because I developed string-based
approaches. An important factor in the success obtained with a program
is to understand precisely its functioning. I thus think that each
user should test different programs, compare them and select the one
that best suits his/her needs.

\section{Requirements}

This part of the tutorial assumes that you already performed the
tutorial about sequence retrieval (above), and that you have the
result files in the current directory. Check with the command:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
cd ${HOME}/practical_rsat
ls -1
\end{verbatim} \end{footnotesize}
}


You should see the following file list:
{\color{OliveGreen} \begin{footnotesize} 
\begin{verbatim}
Escherichia_coli_K12_start_codons.wc
Escherichia_coli_K12_stop_codons.wc
MET_genes.txt
MET_up800-noorf.fasta
PHO_genes.txt
PHO_up800-noorf.fasta
PHO_up800.fasta
RAND_genes.txt
RAND_up800-noorf.fasta
his.genes.txt
his.up200.noorf.fasta
\end{verbatim} \end{footnotesize}
}

\section{oligo-analysis}

The program \program{oligo-analysis} is the simplest pattern discovery
program. It counts the number of occurrences of all oligonucleotides
(words) of a given length (typically 6), and calculates the
statistical significance of each word by comparing its observed and
expected occurrences. The program returns words with a significant
level of over-representation.

Despite its simplicity, this program generally returns good results
for groups of co-regulated genes in microbes.

For a first trial, we will simply use the program to count word
occurrences. The application will be to check the start and stop
codons retrieved above. We will then use \program{oligo-analysis} in a
pattern discovery process, to detect over-represented words from the
set of upstream sequences retrieved above (the PHO family).  In a
first time, we will use the appropriate parameters, which have been
optimized for pattern discovery in yeast upstream sequences (van
Helden et al., 1998). We will then use the sub-optimal settings to
illustrate the fact that the success of word-based pattern-discovery
crucially depends on a rigorous statistical approach (choice of the
background model and of the scoring function).

\subsection{Counting word occurrences and frequencies}

Try the following command:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
oligo-analysis -v 1 -i Escherichia_coli_K12_start_codons.wc \
    -format wc -l 3 -1str
\end{verbatim} \end{footnotesize}
}

Call the on-line option description to understand the meaning of the
options you used:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
oligo-analysis -help
\end{verbatim} \end{footnotesize}
}


Or, to obtain more details:
{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
oligo-analysis -h
\end{verbatim} \end{footnotesize}
}

You can also ask some more information by speifying a verbosity of 1
(option \option{-v 1}), and store the result in a file:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
oligo-analysis -v 1 -i Escherichia_coli_K12_start_codons.wc \
    -format wc -l 3 -1str -return occ,freq \
    -o Escherichia_coli_K12_start_codon_frequencies.tab
\end{verbatim} \end{footnotesize}
}

Read the result file:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
more Escherichia_coli_K12_start_codon_frequencies.tab
\end{verbatim} \end{footnotesize}
}


Note the effect of the verbose option (\option{-v 1}). You receive
information about sequence length, number of possible
oligonucleotides, the content of the output columns, ...


\begin{exercise}
  Follow the same procedure as above to check the frequencies of stop
  codons in the genomes of \org{Escherichia coli K12}, and
  \org{Saccharomyces cerevisia}, respectively.
\end{exercise}

\subsection{Pattern discovery in yeast upstream regions}

Try the following command:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim} 
oligo-analysis -i PHO_up800-noorf.fasta -format fasta    \
    -v 1 -l 6 -2str -lth occ_sig 0 -noov \
    -return occ,proba,rank -sort \
    -bg upstream-noorf -org Saccharomyces_cerevisiae \
    -o PHO_up800-noorf_6nt-2str-noov_ncf_sig0 
\end{verbatim} \end{footnotesize}
}


Note that the return fields (``occ'', ``proba'', and ``rank'') are
separated by a comma \textit{without} space. Call the on-line help to
understand the meaning of the parameters.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim} 
oligo-analysis -h
\end{verbatim} \end{footnotesize}
}


For this analysis, the expected frequency of each word was estimate on
the basis of pre-calibrated frequency tables.  These tables have been
previously calculated (with oligo-analysis) by counting hexanucleotide
frequencies in the whole set of yeast upstream regions (\option{-bg
  upstream}). Our experience is that these frequencies are the optimal
estimator for discovering regulatory elements in upstream sequences of
co-regulated genes.

Analyze the result file:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
more PHO_up800-noorf_6nt-2str-noov_ncf_sig0
\end{verbatim} \end{footnotesize}
}


{\color{OliveGreen} \begin{footnotesize} 
\begin{verbatim}
; Counted on both strands
; 	grouped by pairs of reverse complements
; Background model             	upstream
; Organism                     	Saccharomyces_cerevisiae
; Method                       	Frequency file
...
; Nb of sequences              	19
; Sum of sequence lengths      	11352
; discarded residues           	0	 (other letters than ACGT)
; discarded occurrences        	0	 (contain discarded residues)
; nb possible positions        	11257
; total oligo occurrences      	11257
[...]
; nb possible oligomers        	2080
; oligomers tested for significance	2080
[...]
; column headers
;	1	seq            	oligomer sequence
;	2	identifier     	oligomer identifier
;	3	exp_freq       	expected relative frequency
;	4	occ            	observed occurrences
;	5	exp_occ        	expected occurrences
;	6	occ_P          	occurrence probability (binomial)
;	7	occ_E          	E-value for occurrences (binomial)
;	8	occ_sig        	occurrence significance (binomial)
;	9	rank           	rank
;	10	ovl_occ        	number of overlapping occurrences (discarded from the count)
;	11	forbocc        	forbidden positions (to avoid self-overlap)
;	12	test           	
;seq	identifier	exp_freq	occ	exp_occ	occ_P	occ_E	occ_sig	rank	ovl_occ	forbocc
acgtgc	acgtgc|gcacgt	0.0002182431087	16	2.46	8.4e-09	1.7e-05	4.76	1	2	76
cccacg	cccacg|cgtggg	0.0001528559297	11	1.72	2e-06	4.2e-03	2.37	2	0	55
acgtgg	acgtgg|ccacgt	0.0002257465554	13	2.54	2.8e-06	5.9e-03	2.23	3	1	65
cacgtg	cacgtg|cacgtg	0.0001299168211	10	1.46	3.3e-06	6.8e-03	2.17	4	0	100
cgcacg	cgcacg|cgtgcg	0.0001322750472	10	1.49	3.8e-06	8.0e-03	2.10	5	0	50
cgtata	cgtata|tatacg	0.0005113063008	17	5.76	0.00011	2.2e-01	0.65	6	1	85
agagat	agagat|atctct	0.0006913890231	19	7.78	0.00047	9.8e-01	0.01	7	0	95
\end{verbatim} \end{footnotesize}
}


A few questions:
\begin{enumerate}
\item How many hexanucleotides can be formed with the 4-letter alphabet A,T,G,C ?
\item How many possible oligonucleotides were analysed here ? Is it the
  number you would expect ? Why ?
\item How many patterns have been selected as significant ?
\item By simple visual inspection, can you identify some sequence
  similarities between the selected patterns?
\end{enumerate}

\subsection{Answers}

\begin{enumerate}
\item The number of possible hexanucleotides is $4^6=4,096$.
\item The result file however reports $2,080$ possible
  oligonucleotides. This is due to the fact that the analysis was
  performed on both strands. Each oligonucleotide is thus regrouped
  with its reverse complement. The number of pairs is hovever larger
  than 4096/2, because there are $4^3=64$ motifs (e.g. CACGTG) which
  are identical to their reverse complements. The number of motifs
  distinct from their reverse complement is thus $4,069-64=4,032$, and
  they are regrouped into $4,032/2=2,016$ pairs. The total number of
  motifs is thus $T=64+2016=2080$.
\item Among the 2080 tested oligonucleotides (+reverse complement), no
  more than $7$ were selected as significantly over-represented.
\item Some pairs of words are mutually overlapping
  (e.g. \seq{ACGTGc} and \seq{cACGTG}).
\end{enumerate}


We can now interpret these results in terms of statistics. 

\begin{description}

\item[$exp\_freq$] The expected frequency of an oligonucleotide is the
  probability to find it by chance at any position of the sequences
  analyzed. The expected frequencies are estimated on the basis of the
  background model. 

\item The program \program{oligo-analysis} uses the binomial
  statistics to compare the observed and expected number of
  occurrences, an to calculate the over-representation statistics.

\item[$Pval$] P-value: probability for a given oligonucleotide to be a false
  positive, i.e. to be considered as over-represented whereas it is
  not.

\item[$Eval=T \cdot Pval$] number of false positive patterns expected
  by chance given the P-value of the considered pattern. 

\item[$occ_sig =-log_{10}(Eval)$] significance of the oligonucleotide
  occurrences. This is a simple minus-log conversion of th E-value.
\item
\end{description}


\subsection{Assembling the patterns}

A separate program, \program{pattern-assembly}, allows to assemble a
list of patterns, in order to group those that overlap mutually. Try:

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
pattern-assembly -i PHO_up800-noorf_6nt-2str-noov_ncf_sig0 \
    -v  1 -subst 1 -2str -o PHO_up800-noorf_6nt-2str-noov_ncf_sig0.asmb
\end{verbatim} \end{footnotesize}
}

Read the on-line help to have a look at the assembly parameters. 

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
pattern-assembly -h
\end{verbatim} \end{footnotesize}
}


Let us have a look at the assembled motifs.
 
{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
more PHO_up800-noorf_6nt-2str-noov_ncf_sig0.asmb
\end{verbatim} \end{footnotesize}
}


Should give something llike this (the precise result might be slightly
different depending on the version of the genome).

{\color{OliveGreen} \begin{tiny} 
\begin{verbatim}
; pattern-assembly  -i PHO_up800-noorf_6nt-2str-noov_ncf_sig0 -v 1 -subst 1 -2str -o PHO_up800-noorf_6nt-2str-noov_ncf_sig0.asmb
; Input file    PHO_up800-noorf_6nt-2str-noov_ncf_sig0
; Output file   PHO_up800-noorf_6nt-2str-noov_ncf_sig0.asmb
; Input score column            8
; Output score column           0
; two strand assembly
; max flanking bases            1
; max substitutions             1
; max assembly size             50
; max number of patterns        100
; number of input patterns      7
;

;assembly # 1   seed: acgtgc    9 words length 
;   alignt         rev_cpl      score
cccacg....      ....cgtggg      2.37
cgcacg....      ....cgtgcg      2.10
.gcacgt...      ...acgtgc.      4.76
.ccacgt...      ...acgtgg.      2.23
..cacgtg..      ..cacgtg..      2.17
...acgtgc.      .gcacgt...      4.76
...acgtgg.      .ccacgt...      2.23
....cgtggg      cccacg....      2.37
....cgtgcg      cgcacg....      2.10
cgcacgtgcg      cgcacgtgcg      4.76    best consensus

; Isolated patterns: 2
;alignt rev_cpl score
cgtata  tatacg  0.65    isol
agagat  atctct  0.01    isol
;Job started 26/10/06 09:58:21 CDT
;Job done    26/10/06 09:58:21 CDT
\end{verbatim} \end{tiny}
}

The result of the assembly shows us that several of the significant
hexanucleotides actually reflect various fragments of a same motif.  We
also see that, despite the fact that \program{oligo-analysis} only
analyzed the 4-letters DNA alphabet, the assembly indicates some
degeneracy in the motif, revealed by the presence of alternative
letters at the same position. For instance, in the penultimate
position of the assembly, we can observe either \seq{C} or \seq{G}. In
addition, the scores besides each oligonucleotide indicate us that
these alternative letters can be more or less significantly
over-represented in our sequence set. In summary, the result of
\concept{pattern-assembly} is the real key to the interpretation of
\concept{oligo-analysis}: the discovered motifs are not each separate
oligo-analysis, but the assemblies that can be formed out of them.

The \concept{best consensus} indicates, for each position of the
alignment, the letter corresponding to the oligonucleotide with the
highest significance. This consensus should be considered with
caution, because its complete sequence is built from the collection of
various oligonucleotides, and might not correspond to any real site in
the input sequences. Also, this ``best consensus'' is generally too
stringent to perform pattern matching (see next chapters), and we
usually prefer to search all the oligonucleotides separately, and
analyze their feature map to identify the putative cis-acting
elements.

\begin{exercise}
  Use the same procedure as above to discover over-represented
  hexanucleotides in the upstream sequences of the MET genes obtained
  in the chapter on sequence retrieval. Analyze the results of
  \program{oligo-analysis} and \program{pattern-assembly}.
\end{exercise}


\begin{exercise}
  Use the same procedure as above to discover over-represented
  hexanucleotides in the upstream sequences of the RAND genes (randoms
  election of genes) obtained in the chapter on sequence
  retrieval. Analyze the results of \program{oligo-analysis} and
  \program{pattern-assembly}.
\end{exercise}


\subsection{Alternative background models}

One of the most important parameters for the detectin of significant
motifs is the choice of an appropriate background model. 


This chapter aims at emphasizing how crucial is the choice of
appropriate statistical parameters. We saw above that a background
model calibrated on all the yeast upstream sequences gives good
results with the PHO family: despite the simplicity of the algorithm
(counting non-degenerate hexanucleotide occurrences), we were able to
extract a description of the regulatory motif over a larger width than
6 (by pattern assembly), and we got some description of the degeneracy
(the high and low affinity sites).

We will now intentionally try other parameter settings and see how
they affect the quality of the results.

\subsubsection{Equiprobable oligonucleotides}

Let us try the simplest approach, where each word is considered
equiprobable. For this, we simply suppress the options \option{-bg
  upstream -org Saccharomyces\_cerevisiae} fom the above commands. We
also omit to specify the output file, so results will immediately
appear on the screen.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
oligo-analysis -v 1 -i PHO_up800-noorf.fasta -format fasta \
  -l 6 -2str  -return occ,proba,rank -lth occ_sig 0 -sort -bg equi
\end{verbatim} \end{footnotesize}
}

You can combine \program{oligo-analysis} and
\program{pattern-assembly} in a single command, by using the pipe
character as below.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
oligo-analysis -i PHO_up800-noorf.fasta -format fasta -v 1 \
    -l 6 -2str  -return occ,proba -lth occ_sig 0 -sort \
    | pattern-assembly -2str -subst 1 -v 1 
\end{verbatim} \end{footnotesize}
}

On unix systems, the ``pipe'' character is used to concatenate
commands, i.e. the output of the first command (in this case
\program{oligo-analysis}) is not printed to the screen, but is sent as
input for the second command (in this case \program{pattern-assembly}).

Note that
\begin{itemize} 
\item The number of selected motifs is higher than in the previous
  trial. with the 2006 version of the sequences, I obtain 92 patterns,
  instead of the 7 obtained with the background model calibrated on
  yeast upstream sequences.
\item The most significant motifs are not related to the Pho4p binding
  sites. All these false positives are AT-rich motifs.
\item Two of the selected patterns (\seq{acgttt} and \seq{acgtgc}) are
  related to Pho4p binding site. However, they come at the $56^{th}$ and
  $65^{th}$ positions only. 
\item With this background model, we would thus not be able to detect
  the Pho4p binding sites.
\end{itemize}

\subsubsection{Markov chains}

Another possibility is to use Markov chain models to estimate expected
word frequencies. Try the following commands and compare the
results. None is as good as the option \option{-bg upstream-noorf},
but in case one would not have the pre-calibrated non-coding
frequencies (for instance if the organism has not been completely
sequenced), Markov chains can provide an interesting approach. 

in a Markov chain model, the probability of each oligonucleotide is
estimated on the basis of the probabilities smaller oligonucleotides
that enter in its composition.

We will first apply a Markov model of order 1. 

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
## Markov chain of order 1
oligo-analysis -v 1 -markov 1 \
    -i PHO_up800-noorf.fasta -format fasta \
    -l 6 -lth occ_sig 0 -sort \
    -2str -return occ,proba,rank \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv1

more PHO_up800-noorf_6nt-2str-noov_sig0_mkv1

\end{verbatim} \end{footnotesize}
}

The number of patterns is strongly reduced, compared to the
equiprobable model. A few AT-rich patterns are still rpesent, but the
Pho4p motif now appears at the 3rd position. We can assemble these
oligos in order to highligh the different motifs.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
pattern-assembly -i PHO_up800-noorf_6nt-2str-noov_sig0_mkv1 \
    -2str -sc 7 -subst 1 -v 1 \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv1.asmb

more PHO_up800-noorf_6nt-2str-noov_sig0_mkv1.asmb
\end{verbatim} \end{footnotesize}
}

We can now increase the stringency, by using a Markov model of order 2.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
## Markov chain of order 2
oligo-analysis -v 1 -markov 2 \
    -i PHO_up800-noorf.fasta -format fasta \
    -l 6 -lth occ_sig 0 -sort \
    -2str -return occ,proba,rank \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv2

more PHO_up800-noorf_6nt-2str-noov_sig0_mkv2
\end{verbatim} \end{footnotesize}
}

We now have a very restricted number of patterns, with onnly 2
remaining AT-rich motifs. Besides these, most of the selected oligos
can be assembled to form a moti corresponding to the Pho4p binding
site.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
pattern-assembly -i PHO_up800-noorf_6nt-2str-noov_sig0_mkv2 \
    -2str -sc 7 -subst 1 -v 1 \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv2.asmb

more PHO_up800-noorf_6nt-2str-noov_sig0_mkv2.asmb
\end{verbatim} \end{footnotesize}
}

We can still increase the stringency with a Markov model of order 3.

{\color{Blue} \begin{footnotesize}
\begin{verbatim}
## Markov chain of order 3
oligo-analysis -v 1 -markov 3 \
    -i PHO_up800-noorf.fasta -format fasta \
    -l 6 -lth occ_sig 0 -sort \
    -2str -return occ,proba,rank \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv3

more PHO_up800-noorf_6nt-2str-noov_sig0_mkv3
\end{verbatim} \end{footnotesize}
}

If we further increase the order of the Markov chain, there is not a
single significant oligonucleotide. 

{\color{Blue} \begin{footnotesize}
\begin{verbatim}
## Markov chain of order 4
oligo-analysis -v 1 -markov 4 \
    -i PHO_up800-noorf.fasta -format fasta \
    -l 6 -lth occ_sig 0 -sort \
    -2str -return occ,proba,rank,rank \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv4

more PHO_up800-noorf_6nt-2str-noov_sig0_mkv4
\end{verbatim} \end{footnotesize}
}


\subsubsection{Bernoulli model}

Note that the Markov order 0 means that there is no dependency between
successive residues. The probability of a word is thus simply the
prodct of its residue probabilities. This is a \concept{Bernoulli
  model}, but, by extension of the concepts of Markov chain, it is
accepted to call it markov chain of order 0.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
## Markov chain of order 0 = Bernoulli model
oligo-analysis -v 1 -markov 0 \
    -i PHO_up800-noorf.fasta -format fasta \
    -l 6 -lth occ_sig 0 -sort \
    -2str -return occ,proba,rank \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv0
pattern-assembly -i PHO_up800-noorf_6nt-2str-noov_sig0_mkv0 \
    -2str -sc 7 -subst 1 -v 1 \
    -o PHO_up800-noorf_6nt-2str-noov_sig0_mkv0.asmb

more PHO_up800-noorf_6nt-2str-noov_sig0_mkv0.asmb
\end{verbatim} \end{footnotesize}
}


\subsubsection{Summary about the Markov chain background models}

\begin{itemize}
\item The Markov model of order 1 returns AT-rich patterns with the
  highest significance, but the Pho4p high affinity site is described
  with a good accuracy. The medium affinity site appears as a single
  word (acgttt) in the isolated patterns.
\item Markov order 1 returns less AT-rich motifs. The poly-A (aaaaaa)
  is however still associated with the highest significance, but comes
  as isolated pattern.
\item The higher the order of the markov chain, the most stringent are
  the conditions. For small sequence sets, selecting a too high order
  prevents from selecting any pattern. Most of the patterns are missed
  with a Markov chain of order 2, and higher orders don't return any
  single significant word.
\end{itemize}


\section{Genome-scale pattern discovery}

The detection of exceptional words can also be used to detect signals
in large sequence sets, such as th whole set of upstream sequences for
a given organism, or even its complete genome. We will illustrate this
with two examples.

\subsection{Detection of over-represented words in all the yeast
  upstream   sequences}

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
retrieve-seq -org Saccharomyces_cerevisiae -type upstream -all \
    -from -1 -to -800 -noorf -o Saccharomyces_cerevisiae_allup_800-noorf.fasta.gz
\end{verbatim} \end{footnotesize}
}

Note that we added the extension \texttt{.gz} to the name of the
output file. This suffix is interpreted by all the \RSAT programs as
an indication to compress the result using the command
\command{gzip}. The result file occupies a much smaller space on your
hard drive.

We will now analyze the frequency of all the heptanucleotides, and
analyze their level of over- or under-representation (for this, we use
the option \option{-two\_tails}). To estimate expected frequencies, we
will use a Markov model of order 4 (the other models are left as
exercise).

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
oligo-analysis -v 1 -i Saccharomyces_cerevisiae_allup_800-noorf.fasta.gz \
    -l 7 -2str -noov -return occ,freq,proba,zscore,rank -sort -markov 4 \
    -two_tails \
    -o   Saccharomyces_cerevisiae_allup_800-noorf_7nt-2str-noov_mkv4.tab
\end{verbatim} \end{footnotesize}
}

you can now compare the most significant oligonucleotides with the
transcription factor binding sites annotated in SCPD, the Sacharomyces
cerevisiae Promoter Database
(\url{http://rulai.cshl.edu/cgi-bin/SCPD/searchmotif}).

\subsection{Detection of under-represented words in bacterial genomes}

\begin{exercise}

  Analyze the frequencies of al the hexanucleotides in
  \org{Escherichia coli K12}. One of them shows a very high degree of
  under-representation. Try to understand the reason why this
  hexanucleotide is avoided in this genome.

  \textbf{Info:} the full genome of \org{Escherichia coli K12} can be
  found in the \RSAT genome directory.

{\color{Blue} \begin{footnotesize} 
\begin{verbatim}
ls $RSAT/data/genomes/Escherichia_coli_K12/genome/contigs.txt
\end{verbatim} \end{footnotesize}
}

This file contains the list of chromosomes of the bacteria (in this
cxase there is a single one, but for \org{S.cerevisiae} there are 16 nuclear
and one mitochondrial chromosomes). It can be idrectly used as input
by specifying the format \option{-format filelist}.
\end{exercise}

% {\color{Blue} \begin{footnotesize} 
% \begin{verbatim}
% oligo-analysis -v 1 -format filelist -i $RSAT/data/genomes/Escherichia_coli_K12/genome/contigs.txt -return occ,proba,zscore,rank -under -l 6 -sort -o Escherichia_coli_K12_genome_6nt-2str-noov_mkv0_under -2str -noov -markov 0
% \end{verbatim} \end{footnotesize}
% }

\section{dyad-analysis}

In the previous chapter, we saw that \program{oligo-analysis} allows
to detect over- and under-represented motifs in biological sequences,
according to a user-specified background model. Since 1997, this
program has been routinely used to predict cis-acting elements from
groups of co-expressed genes.

However, some motifs escape to \program{oligo-analysis}, because they
do not correspond to an oligonucleotide, but to a spaced pair of very
short oligonucleotides (\concept{dyads}). To address this problem, we
developed another program called \program{dyad-analysis}.

\tbw

%% \section{gibbs motif sampler (program developed by Andrew Neuwald)}



