%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sequence retrieval
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Retrieving sequences}

The program \program{retrieve-seq} allows you to retrieve sequences
from a genome (provided this genome is supported on your machine). In
particular (and by default), this program extracts the non-coding
sequences located upstream the start codon of the query genes. The
reason for selecting upstream sequences (rather than coding) is that
regulatory elements are generally found upstream of the coding
regions, at least in microbial organisms.

\section{Retrieving a single upstream sequence}

First trial: we will extract the upstream sequence for a single
gene. Try:

\begin{lstlisting}
retrieve-seq -type upstream -org Escherichia_coli_K12 \
    -q metA -from -200 -to -1
\end{lstlisting}


This command retrieves a 200 bp upstream sequence for the gene
\gene{metA} of the bacteria \org{Escherichia coli K12}. 

By default, coordinates are calculated from the start codon. Ideally,
we would prefer to retrieve sequences upstream of the Transcription
Start Site (\concept{TSS}), since this is the place where the RNA
polymerase starts to transcribe the gene. Unfortunately, the precise
location of the TSS is unknown for most genes, in most sequecned
genome. For this reason, the default reference is the start codon
rather than the TSS.

Note that for some organisms (e.g. \org{Homo sapiens}), genome
annotations include mRNA boundaries. In this case, the option
\option{-feattype mRNA} allows you to specify that the reference point
is the start of the mRNA (thus the TSS) rather than the start codon.

Whichever reference point you decide to use, negative coordinates
indicate sequences upstream to this reference point, and positive
coordinates downstream sequences. 

\begin{samepage}
With the default parameters, 
\begin{itemize}
\item[-] the reference point is the start codon;
\item[-] position $-1$ corresponds to the first residue upstream of
  the coding sequence; 
\item[-] position 0 is the first letter from the start codon (the A from
  ATG);
\item[-] positive coordinates indicate the coding sequence (downstream
  from the start codon).
\end{itemize}
\end{samepage}

To better understand the system of coordinates, try to locate the
start codon in the sequence obtained with the following commands.

\begin{lstlisting}
retrieve-seq -type upstream -org Escherichia_coli_K12 \
    -q metA -from -5 -to 6
\end{lstlisting}


\section{Combining upstream and coding sequence}

For \org{E.coli} genes, regulatory signals sometimes overlap the 5'
side of the coding sequence. By doing so, they exert a repression
effect by preventing RNA-polymerase from binding DNA. The command
\program{retrieve-seq} allows you to extract a sequence that overlaps
the start codon, to combine an upstream and a coding segment.

\begin{lstlisting}
retrieve-seq -type upstream -org Escherichia_coli_K12 \
    -q metA -from -200 -to 49
\end{lstlisting}

\section{Retrieving a few upstream sequences}

The option \option{-q} (query gene) can be used iteratively in a
command to retrieve sequences for several genes.

\begin{lstlisting}
retrieve-seq -org Escherichia_coli_K12 \
    -from -200 -to 49 -q metA -q metB -q metC
\end{lstlisting}

\section{Retrieving a larger list of upstream sequences}

If you have to retrieve a large number of sequences, it might become
cumbersome to type each gene name on the command-line. A list of gene
names can be provided in a text file, each gene name coming as the
first word of a new line.

As an example, we willuse the command \command{gene-info} to collect
all genes whose matches the prefix \texttt{PHO} followed by one or
several numbers (this will return a list of genes involved in
phosphate metabolism).

\begin{lstlisting}
gene-info -org Saccharomyces_cerevisiae -q 'PHO\d+' -o PHO_genes.txt
\end{lstlisting}

We can check the content of your file by typing

\begin{lstlisting}
cat PHO_genes.txt
\end{lstlisting}

This file can now be used as input to indicate the list of query genes
for \command{retrieve-seq}, with the option \option{-i}.

\begin{lstlisting}
retrieve-seq -type upstream -i PHO_genes.txt \
    -org Saccharomyces_cerevisiae \
    -from -800 -to -1  
\end{lstlisting}

The option \option{-o} allows you to indicate the name of a file where
the sequence will be stored.

\begin{lstlisting}
retrieve-seq -type upstream -i PHO_genes.txt \
    -org Saccharomyces_cerevisiae \
    -from -800 -to -1   -label name \
    -o PHO_up800.fasta
\end{lstlisting}

Check the sequence file:

\begin{lstlisting}
more PHO_up800.fasta
\end{lstlisting}


\section{Preventing the inclusion of upstream ORFs}

With the command above, we retrieved sequences covering precisely 200
bp upstream the start codon of the selected genes. Intergenic regions
are sometimes shorter than this size. In particular, in bacteria, many
genes are organized in operons, and the intergenic distance is very
short (typically between 0 and 50 bp). If your gene selection contains
many intra-operon genes, the sequences will be mainly composed of
coding sequences (more precisely ORF, open reading frame), which will
bias subsequent analyses.

The option \option{-noorf} of \textit{retrieve-seq} indicates that, if
the upstream gene is closer than the specified limit, the sequence
should be clipped in order to return only intergenic regions.

As an example, we will store the list of histidin genes in a file and
compare the results obtained with and without the option
\option{-noorf}.

Create a text file named \file{his\_genes.txt} with the following
genes.

\begin{footnotesize}
\begin{verbatim}
hisL
hisG
hisD
hisC
hisH
hisA
hisF
hisI
hisP
hisM
hisQ
hisJ
hisS
\end{verbatim}
\end{footnotesize}

The default behaviour will return 200bp for each gene. 

\begin{lstlisting}
retrieve-seq -type upstream -org Escherichia_coli_K12 \
    -i his_genes.txt -from -200 -to -1
\end{lstlisting}

With the option \option{-noorf}, sequences are clipped depending on
the position of the closest upstream neighbour.

\begin{lstlisting}
retrieve-seq -type upstream -org Escherichia_coli_K12 \
    -i his_genes.txt -from -200 -to -1 -noorf \
    -o his_up200_noorf.fasta

more his_up200_noorf.fasta
\end{lstlisting}

You can measure the length of the resulting sequences with the program
\program{sequence-lengths}.

\begin{lstlisting}
sequence-lengths -i his_up200_noorf.fasta
\end{lstlisting}

Notice that some genes have very short upstream sequences (no more
than a few bp, or even 0bp). These are the internal genes of the
\gene{his} operon. 


We will now apply the same option to the list of PHO genes entered
above, in order to obtaine the corresponding non-coding upstream
sequences, with a size up to 800bp.

\begin{lstlisting}
retrieve-seq -type upstream -i PHO_genes.txt \
    -org Saccharomyces_cerevisiae \
    -from -800 -to -1 -noorf  -label name \
    -o PHO_up800-noorf.fasta
\end{lstlisting}

Check the sequence file:

\begin{lstlisting}
more PHO_up800-noorf.fasta
\end{lstlisting}

We can now use the command \command{sequence-lengths} to compare the
sequence sizes of the files \file{PHO\_up800.fasta}, and
\file{PHO\_up800-noorf.fasta}, respectively.

\begin{lstlisting}
sequence-lengths -i PHO_up800.fasta

sequence-lengths -i PHO_up800-noorf.fasta
\end{lstlisting}

\section{Getting information about genes}

\RSAT include several utilities to obtain information about a set of
genes, we will illustrate some basic features. 

\subsection{Getting gene location, names and description}

In the previous section, we created a text file with the names of a
set of genes related to phosphate metabolism. The command
\command{gene-info} returns the complete information concerning a set
of genes. By default, the first word of each row of the input file is
considered as a query.

\begin{lstlisting}
gene-info -i PHO_genes.txt -org Saccharomyces_cerevisiae
\end{lstlisting}


\subsection{Selecting gene by name or description}

Another common need is to search all the names whose name or
description matches some string. For example, let us assume that we
want to ollect all the genes whose name indicates a role in the
methionine metabolism, in the yeast \org{Saccharomyces cerevisiae}.
The program \concept{gene-info} allows us to specify this type of
query. according to the naming convention in the yeast community, gene
names start with three letters indicating the function (e.g. PHO for
phosphate, MET for methionine), wollowed by a number. We can ask the
program to return all the gene names having the string ``MET'' in
their names. 

In this example, we will enter the query string with the option
\option{-q} on the command line, rather than in a file.

\begin{lstlisting}
gene-info -q 'MET' -org Saccharomyces_cerevisiae
\end{lstlisting}

We could also refine the query by taking advantage of our knowledge of
the yeast gene nomenclature, and selecting the genes whose name starts
with the prefix ``MET'', followed by one or several numbers.

\begin{lstlisting}
gene-info -q '^MET\d+' -org Saccharomyces_cerevisiae
\end{lstlisting}

The query is formulatd as a \concept{regular expression}, where
\texttt{$\backslash$d} indicates a number, and the symbol $+$ is a
multiplier, so \texttt{$\backslash$d+}, indicates that we accept a
succession of one or more numbers after the string ``MET''. The
character $\hat{ }$ indicates that the string MET should be at the start of
the name (thus, there can be no letter before MET).

We can now store this list of genes in a separate file, and retrieve
the coresponding upstream sequences.


\begin{lstlisting}
gene-info -q '^MET\d+' -org Saccharomyces_cerevisiae -o MET_genes.txt

retrieve-seq -type upstream -i MET_genes.txt \
    -org Saccharomyces_cerevisiae \
    -from -800 -to -1 -noorf  -label name \
    -o MET_up800-noorf.fasta
\end{lstlisting}


\subsection{Selecting genes by their description}

By default, the program \concept{gene-info} matches a query string
against the list of gene names for the selected organism. The option
\option{-descr} extends the search to the gene descriptions. For
instance, we could search all the genes having the word ``methionine''
in their description. 

\begin{lstlisting}
gene-info -descr -q methionine -org Saccharomyces_cerevisiae
\end{lstlisting}

\subsection{Adding selected fields to a list of gene}

As we saw in the previous section, the program \program{gene-info}
takes as input a list of gene names or identifiers, and return the
complete description of each gene.

In some cases, one needs only a part of this information (e.g. the
common name, or the descripion), in order to to add some columns to a
pre-existing tab-delimited file where each row represents one
gene. For example, imagine that you have a file containing expression
profiles for 6,000 yeast genes, measured by microarray experiments
under 200 conditions. The file contains 201 columns: the first column
indicates the ID of each gene, and the 200 next column give expression
values measured in the 200 microarrays. In such case, you would
typically use \program{add-gene-info} to add a few columns after each
profile, in order to indicate the common name and the description of
each gene.

The program \concept{add-gene-info} allows add columns to an input
file, with user-selected fields of information about the genes. For
example, the options below will add the gene identifier and the list
of synonym to each row of our PHO gene list. 

\begin{lstlisting}
add-gene-info -i PHO_genes.txt -org Saccharomyces_cerevisiae \
  -info id,names
\end{lstlisting}

If the input file contains additional columns (e.g. expression
profiles), these will be preserved in the output, and the requested
information columns will be added at the end of each row.

You can check the list of fields supported by \program{add-gene-info}
by consulting the help message.

\begin{lstlisting}
add-gene-info -help
\end{lstlisting}


\section{Retrieving sequences of a random selection of genes}

It is also sometimes interesting to select a set of random genes,
which canbe used as negative control or some analyses. This is exactly
the purporse of the program \program{random-genes}. We will perform a
random selection of 20 yeast genes, and retrieve their upstream
sequences. This selection will also be used in the next chapters.


\begin{lstlisting}
random-genes -org Saccharomyces_cerevisiae -n 20 -o RAND_genes.txt

retrieve-seq -type upstream -i RAND_genes.txt \
    -org Saccharomyces_cerevisiae \
    -from -800 -to -1 -noorf  -label name \
    -o RAND_up800-noorf.fasta
\end{lstlisting}

\section{Retrieving all upstream sequences}

For genome-scale analyses, it is convenient to retrieve upstream
sequences for all the genes of a given genome, without having to
specify the complete list of names. For this, simply use the option
\option{-all}.

As an illustration, we will use \command{retrieve-seq} to retrieve all
the start codons from \org{Escherichia coli}. As we saw before,
negative coordinates specify upstream positions, 0 being the first
base of the coding sequence. Thus, by specifying positions 0 to 2, we
will extract the three first coding bases, i.e. the start codon. 

\begin{lstlisting}
retrieve-seq -type upstream -org Escherichia_coli_K12 \
    -from 0 -to 2 \
    -all -format wc -nocomments -label id,name \
    -o Escherichia_coli_K12_start_codons.wc
\end{lstlisting}

Check the result:

\begin{lstlisting}
more Escherichia_coli_K12_start_codons.wc
\end{lstlisting}


\section{Retrieving downstream sequences}

\program{retrieve-seq} can also be used to retrieve downstream
sequences. In this case, the origin (position 0) is the third base of
the stop codon, positive coordinates indicate downstream (3')
location, and negative coordinates locations upstream (5') from the
stop codon (i.e. coding sequences). 

For example, the following command will retrieve 200pb downstream
sequences for a few yeast genes. The first nucleotides of the
retrieved sequences are those immediately after the stop codon.

\begin{lstlisting}
retrieve-seq -type downstream -org Saccharomyces_cerevisiae \
    -from 1 -to 200 -label id,name -q PHO5 -q MET4
\end{lstlisting}

Since with the option \option{-type downstream}, the coordinates
smaller than 1 indicate positions upstream of the stop codon, we can
use \program{retrieve-seq} to extract the stop codons for all the
genes of \org{Escherichia coli}.

\begin{lstlisting}
retrieve-seq -type downstream -org Escherichia_coli_K12 \
    -from -2 -to 0 \
    -all -format wc -nocomments -label id,name \
    -o Escherichia_coli_K12_stop_codons.wc
\end{lstlisting}

\section{Inferring operons}

In Bacteria, genes are organized in operon, which means that several
genes are transcribed in a single transcription unit. The
transcription of a whole operon is driven by a single promoter,
located upstream of the so-called \concept{leader gene}.

Let us assume that we dispose of a set of bacterial genes for which we
want to predict cis-acting elements (e.g. co-expressed genes in a
microarray experiment). A good fraction of these genes might be
located inside operons. For these, the putative regulatory elements
should be searched in the promoter of the operon leader gene, rather
than in the upstream sequence of the gene itself.

The program \program{infer-operon} allows to infer the operons and
return the corresponding leader genes for a set of input genes.  The
approach is inspired by the Salgado-Hagelsieb method, which consists
in predicting, for each upstream region, if it is within an operon
(WO) or t a transcription unit border (TUB). This prediction is based
on two rules:

\begin{enumerate}
\item \textbf{Orientation rule} If the intergenic region is flanked by two
  genes located on different strands, it is a TUB.

\item \textbf{[Distance rule} If the intergenic region is flanked two
  \concept{tandem} genes (adjacent genes transcribed in the same
  direction), it is classified as WO if the intergenic distance is
  lower than some threshold (by default, 55bp), and as TUB otherwise.
\end{enumerate}

The default distance threshold was chosen to obtain a good balance
between \concept{sensitivity} (\concept{Sn}, fraction of annotated WO
regions which are correctly predicted) and \concept{positive
  predictive value} (\concept{PPV}, fraction of predicted WO region
which indeed correspond to annotations).

The option \option{-dist} allows to specify a custom distance
threshold. By increasing the threshold, the number of regions
predicted as WO increases, at the expense of those predicted as
WO. This will thus increase the Sn and decrease PPV.

The \concept{accuracy} measures the balance between Sn and PPV by
taking their arithmetic average. With the default value, one can
expect ~78\% of accuracy (Reki's janky and Jacques van Helden,
unpublished results).

We will illustrate the use of \program{infer-operons} with a few
examples.
    
\subsection{Inferring operon from a list of query genes}

With the following command, we infer the operon for a set of input
genes.

\begin{lstlisting}
infer-operon -v 1 -org Escherichia_coli_K12 -q hisD -q mhpR \
  -q mhpA -q mhpD
\end{lstlisting}

Note that the prediction is incorrect for the gene \gene{hisD}: the
program predict \gene{hisG} as operon leader, whereas the well known
leader of the \gene{his} operon is \gene{hisL}. This is due to the
fact that the intergenic distance between \gene{hisL} and \gene{hisG}
is 145bp, which exceeds the default distance threshold (55bp). 

One option would be to increase the distance threshold to 150bp. 


\begin{lstlisting}
infer-operon -v 1 -org Escherichia_coli_K12 -q hisD -q mhpR \
  -q mhpA -q mhpD -dist 150
\end{lstlisting}

However, we should be very careful with this option, since it has a
strong consequence on all the other operon inferenes in the same
genome. Since a good fraction of promoters of \org{Escherichia coli}
are shorted than 150bp, by increasing the distance threshold to 150,
we will undully consider these promoters as WO.

\subsection{Selecting custom return fields}

The option \option{-return} allows to specify custom return fields.

\begin{lstlisting}
infer-operon -v 1 -org Escherichia_coli_K12 -q hisD -q lacI -q lacZ \
  -return q_info,up_info,leader,trailer,operon
\end{lstlisting}

Note that the famous \gene{lac} operon contains three genes:
\gene{lacZ}, \gene{lacY} and \gene{lacA}, but the inferred operon only
returns the two first genes because the distance between \gene{lacY}
and \gene{lacA} is 65bp. This can be checked with the return field
\option{down\_info}.

\begin{lstlisting}
infer-operon -v 1 -org Escherichia_coli_K12 -q lacZ -q lacY \
  -return q_info,up_info,down_info,operon
\end{lstlisting}

\subsection{Operons with non-CDS genes}

Note that operons can contain non-coding genes. For example, the
metT operon contains a series of tRNA genes for methionine, leucine
and glutamina, respectively.

\begin{lstlisting}
infer-operon -org Escherichia_coli_K12 -q glnV -q metU -q ileV \
-return q_info,up_info,operon
\end{lstlisting}

\subsection{Inferring all operons for a given organism}

The option \option{-all} allows to infer operons for all the genes of
an organism.

\begin{lstlisting}
infer-operon -v 1 -org Escherichia_coli_K12 -all \
  -return q_info,up_info,leader,operon
\end{lstlisting}

\subsection{Retrieving operon leader genes and inferred
  operon promoters}

As explained above, a common usage of operon inference is to predict a
list of leader genes from a set of query genes, in order to retrieve
the corresponding promoter sequences. For this, we will use the option
\option{-return} to obtain the leader gene in the first column of the
result table. 

\begin{lstlisting}
infer-operon -org Escherichia_coli_K12 -return leader,q_info,up_info,operon \
   -q lacI -q lacZ -q lacY -q mhpD -q mhpF
\end{lstlisting}

The first column now indicates the inferred leader genes rather than
tne query genes, and that this column contains some redundancy: the
same leader gene appears multiple times. This comes from the fact that
several of our query genes were part of the same operon (e.g.:lacZ and
lacY).

To avoid including twice their leader, we use the unix command sort -u
(unique).


\begin{lstlisting}
infer-operon -org Escherichia_coli_K12 -return leader,q_info,up_info,operon \
   -q lacI -q lacZ -q lacY -q mhpD -q mhpF \
   | cut -f 1 \
   | sort -u 
\end{lstlisting}

We can now use the resulting non-redundant list of operon leaders as
input for retrieve-seq.

\begin{lstlisting}
infer-operon -org Escherichia_coli_K12 -return leader,q_info,up_info,operon \
   -q lacI -q lacZ -q lacY -q mhpD -q mhpF \
   | cut -f 1 \
   | sort -u \
   | retrieve-seq -org Escherichia_coli_K12 -noorf
\end{lstlisting}


\subsection{Collecting all upstream regions from the query gene up to
  the leader gene}

\textbf{TO BE IMPLEMENTED}

\subsection{Automatic inference}

\textbf{TO BE IMPLEMENTED}


\section{Purging sequences}

\tbw
